---
title: "Assignment 04"
authors: Keegan Brown, Patrick Jones
format: html
editor: visual
  self-contained: true
editor_options: 
  chunk_output_type: console
---

#Loading packages and data

```{r}
library(tidyverse)
library(haven)
library(sf)
library(tigris)


#loading 
data <- read_xpt("LLCP2021.XPT ")

```

Paragraph describing the data here

#Visualization 1 ##reviewing employment status and ACE scores Adverse childhood experiences are strong predictors of bad health outcomes, increased incarceration rate, and lower education outcomes. The greater number of reported ACE experiences are associated with worse employment outcomes. Using a (logit/probit model, reminder to review for cat variable), we assess the impact of increased ACE experiences on the likelihood that an indivdual is unemployed.

### Recoding Employment and ACE Data

```{r}

### employment variable 
#EMPLOY1 -
## 1 = employed 4 waged
## 2 = self employed
## 3 = unemployed 1+ year
## 4 = unemployed less than 1 year
## 5 = homemaker - coding as working
## 6 = student - coding as working
## 7 = retired - dropping from dataset 
## 8 = unable to work - coding as unemployed
## 9 = refused - dropping from dataset 
## n/a = missing - dropping from dataset 
### ace variables 

g1data <- data%>%
  mutate(
    employ_dum = case_when(
      EMPLOY1 %in% c(1,2,5,6) ~ "1",
      EMPLOY1 %in% c(3,4,8) ~ "0"
    )
  )

g1data <- g1data%>%
  filter(!(EMPLOY1 == 9), 
         !(EMPLOY1 == 7),
         !(is.na(EMPLOY1)))


### ACE Variables 
##ACEDEPRS 
#1 = yes 
#2 = no
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_deprs = case_when(
      ACEDEPRS %in% c(1) ~ "1",
      ACEDEPRS %in% c(2,7) ~ "0",
      is.na(ACEDEPRS) ~ "0"
    )
  )


#ACEDRINK 
#1 = yes 
#2 = no
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_drink = case_when(
      ACEDRINK %in% c(1) ~ "1",
      ACEDRINK %in% c(2,7) ~ "0",
      is.na(ACEDRINK) ~ "0"
    )
  )

#ACEDRUGS
#1 = yes 
#2 = no
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_drug= case_when(
      ACEDRUGS %in% c(1) ~ "1",
      ACEDRUGS %in% c(2,7) ~ "0",
      is.na(ACEDRUGS) ~ "0"
    )
  )


#ACEPRISN 
#1 = yes 
#2 = no
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_prisn = case_when(
      ACEPRISN %in% c(1) ~ "1",
      ACEPRISN %in% c(2,7) ~ "0",
      is.na(ACEPRISN) ~ "0"
    )
  )


#ACEDIVRC 
#1 = yes 
#2 = no
#7 = don't know not sure - coding as no
#8 = parents not married - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_div = case_when(
      ACEDIVRC %in% c(1) ~ "1",
      ACEDIVRC %in% c(2,7,8) ~ "0",
      is.na(ACEDIVRC) ~ "0"
    )
  )


#ACEPUNCH
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_punch = case_when(
      ACEPUNCH %in% c(1,7) ~ "0",
      ACEPUNCH %in% c(2,3) ~ "1",
      is.na(ACEPUNCH) ~ "0"
    )
  )

#ACEHURT1 
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_hurt = case_when(
      ACEHURT1 %in% c(1,7) ~ "0",
      ACEHURT1 %in% c(2,3) ~ "1",
      is.na(ACEHURT1) ~ "0"
    )
  )


#ACESWEAR
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping


g1data <- g1data%>%
  mutate(
    ace_swear = case_when(
      ACESWEAR %in% c(1,7) ~ "0",
      ACESWEAR %in% c(2,3) ~ "1",
      is.na(ACESWEAR) ~ "0"
    )
  )

#ACETOUCH 
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_touch = case_when(
      ACETOUCH %in% c(1,7) ~ "0",
      ACETOUCH %in% c(2,3) ~ "1",
      is.na(ACETOUCH) ~ "0"
    )
  )

#ACETTHEM
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_tthem = case_when(
      ACETTHEM %in% c(1,7) ~ "0",
      ACETTHEM %in% c(2,3) ~ "1",
      is.na(ACETTHEM) ~ "0"
    )
  )

#ACEHVSEX 
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_hvsex = case_when(
      ACEHVSEX %in% c(1,7) ~ "0",
      ACEHVSEX %in% c(2,3) ~ "1",
      is.na(ACEHVSEX) ~ "0"
    )
  )

#ACEADSAF 
#1 = never 
#2 = a little of the timne - coding as yes
#3 = some of the time - coding as yes
#4 - most of the time - coding as yes
#5 - all of the time - coding as yes
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_adsaf = case_when(
      ACEADSAF %in% c(1,7) ~ "0",
      ACEADSAF %in% c(2,3,4,5) ~ "1",
      is.na(ACEADSAF) ~ "0"
    )
  )


#ACEADNED 
#1 = never 
#2 = a little of the timne - coding as yes
#3 = some of the time - coding as yes
#4 - most of the time - coding as yes
#5 - all of the time - coding as yes
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_adned = case_when(
      ACEADNED %in% c(1,7) ~ "0",
      ACEADNED %in% c(2,3,4,5) ~ "1",
      is.na(ACEADNED) ~ "0"
    )
  )

## drops 

g1data <- g1data%>%
  filter(ace_deprs != 9,
         ace_drink != 9,
         ace_drug != 9,
         ace_prisn != 9, 
         ace_div != 9,
         ace_punch !=9,
         ace_hurt != 9,
         ace_swear != 9,
         ace_touch != 9, 
         ace_tthem != 9,
         ace_hvsex != 9,
         ace_adsaf != 9,
         ace_adned != 9
  )

###Ace count score 
acecount <- g1data%>%
  

ggplot(data = g1data)+
  geom_line(mapping = aes())+
  geom_point() 
  


```

###test code block

```{r}


     
```

#Visualization 2

```{r}

```

#Visualization 3

```{r}
g3data <- data

g3data %>%
  filter(PHYSHLTH <= 30) %>%
  filter(INCOME3 <= 11) %>%
  filter(`_AGEG5YR` <= 13) %>%
  ggplot(mapping = aes(x = `_AGEG5YR`, y = INCOME3, fill = PHYSHLTH)) +
geom_tile() +
  coord_cartesian(xlim = c(1, 13), ylim = c(1, 11)) +
  scale_fill_continuous(
    low = "#1097b8", 
  high = "#b83110", 
  name = "Days of poor physical\nhealth per month") +
  scale_y_continuous(breaks = c(1:11),
                   labels = c("Less than $10,000",
                              "$10,000 - $15,000",
                              "$15,000 - $20,000",
                              "$20,000 - $25,000",
                              "$25,000 - $35,000",
                              "$35,000 - $50,000",
                              "$50,000 - $75,000",
                              "$75,000 - $100,000",
                              "$100,000 - $150,000",
                              "$150,000 - $200,000",
                              "$200,000 or more")) +
  scale_x_continuous(breaks = c(1:13),
                     labels = c("18 - 24",
                                "25 - 29",
                                "30 - 34",
                                "35 - 39",
                                "40 - 44",
                                "45 - 49",
                                "50 - 54",
                                "55 - 59",
                                "60 - 64",
                                "65 - 69",
                                "70 - 74",
                                "75 - 79",
                                "80 or older")) +
  labs(title = "Physical health among age and income groups, U.S. adults, 2021",
       subtitle = "Young, wealthy Americans report highest levels of\nphysical health, while poor, elderly Americans report the lowest",
       caption = "Behavioral Risk Factor Surveillance System, 2021") +
  xlab("Age") +
  ylab("Income") +
  theme_minimal() +
  theme(text = element_text(family = "Helvetica"),
        axis.text.x = element_text(
          angle = 45,
          vjust = 1,
          hjust = 1
        ),
        axis.title.y = element_text(
          angle = 0,
          hjust = 0,
          vjust = 1
        ),
        axis.title.x = element_text(
          hjust = .5,
          vjust = 1
        ))
  
```

#Visualization 4

```{r}
g4data <- data

g4data <- g4data %>%
  filter(MENTHLTH <= 30) %>%
  group_by(INCOME3) %>%
  summarize(mean_MENTHLTH = mean(MENTHLTH))

g4data %>%
  filter(INCOME3 < 77) %>%
  ggplot(aes(x = INCOME3, 
             y = mean_MENTHLTH,
             fill = mean_MENTHLTH)) +
  geom_col() +
  scale_fill_continuous(low = "#1097b8", 
                        high = "#b83110",
                        guide = "none") +
  scale_x_continuous(breaks = c(1:11),
                   labels = c("Less than $10,000",
                              "$10,000 - $15,000",
                              "$15,000 - $20,000",
                              "$20,000 - $25,000",
                              "$25,000 - $35,000",
                              "$35,000 - $50,000",
                              "$50,000 - $75,000",
                              "$75,000 - $100,000",
                              "$100,000 - $150,000",
                              "$150,000 - $200,000",
                              "$200,000 or more")) +
  coord_flip() +
  labs(title = "Mental health by income groups, U.S. adults, 2021",
       subtitle = "Poor Americans report more days of mental distress than wealthy Americans",
       caption = "Behavioral Risk Factor Surveillance System, 2021",) +
  xlab("Annual income") +
  ylab("Average days of mental distress per month") +
  theme_minimal() +
  theme(text = element_text(family = "Helvetica"),
        axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                     vjust = 1,
                                    )) 
  

```
