---
title: "Assignment 04"
authors: Keegan Brown, Patrick Jones
format: html
editor: visual
editor_options: 
  chunk_output_type: console
output:
  html_document:
    self_contained: true
---

#Loading packages and data

```{r}
library(tidyverse)
library(haven)
library(sf)
library(tigris)
library(tidycensus)
library(RColorBrewer)



#loading 
data <- read_xpt("LLCP2021.XPT ")

```

Paragraph describing the data here

#Visualization 1 
This line graph plots employment status with the quantity of reported Adverse Childhood Experiences (ACE). ACEs are strong predictors of poor health outcomes, increased incarceration rate, and lower education outcomes. Here, the unemployment is shown to sharply rise as the ACE score increases, suggesting appropriate intervention at a young age can have positive benefits for future labor market health. 



```{r}

### employment variable 
#EMPLOY1 -
## 1 = employed 4 waged
## 2 = self employed
## 3 = unemployed 1+ year
## 4 = unemployed less than 1 year
## 5 = homemaker - coding as working
## 6 = student - coding as working
## 7 = retired - dropping from dataset 
## 8 = unable to work - coding as unemployed
## 9 = refused - dropping from dataset 
## n/a = missing - dropping from dataset 
### ace variables 

g1data <- data%>%
  mutate(
    employ_dum = as.numeric(
      case_when(
      EMPLOY1 %in% c(1,2,5,6) ~ "1",
      EMPLOY1 %in% c(3,4,8) ~ "0"
    )
  )
)

g1data <- g1data%>%
  filter(!(EMPLOY1 == 9), 
         !(EMPLOY1 == 7),
         !(is.na(EMPLOY1)))


### ACE Variables 
##ACEDEPRS 
#1 = yes 
#2 = no
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_deprs = case_when(
      ACEDEPRS %in% c(1) ~ "1",
      ACEDEPRS %in% c(2,7) ~ "0",
      is.na(ACEDEPRS) ~ "0"
    )
  )


#ACEDRINK 
#1 = yes 
#2 = no
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_drink = case_when(
      ACEDRINK %in% c(1) ~ "1",
      ACEDRINK %in% c(2,7) ~ "0",
      is.na(ACEDRINK) ~ "0"
    )
  )

#ACEDRUGS
#1 = yes 
#2 = no
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_drug= case_when(
      ACEDRUGS %in% c(1) ~ "1",
      ACEDRUGS %in% c(2,7) ~ "0",
      is.na(ACEDRUGS) ~ "0"
    )
  )


#ACEPRISN 
#1 = yes 
#2 = no
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_prisn = case_when(
      ACEPRISN %in% c(1) ~ "1",
      ACEPRISN %in% c(2,7) ~ "0",
      is.na(ACEPRISN) ~ "0"
    )
  )


#ACEDIVRC 
#1 = yes 
#2 = no
#7 = don't know not sure - coding as no
#8 = parents not married - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_div = case_when(
      ACEDIVRC %in% c(1) ~ "1",
      ACEDIVRC %in% c(2,7,8) ~ "0",
      is.na(ACEDIVRC) ~ "0"
    )
  )


#ACEPUNCH
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_punch = case_when(
      ACEPUNCH %in% c(1,7) ~ "0",
      ACEPUNCH %in% c(2,3) ~ "1",
      is.na(ACEPUNCH) ~ "0"
    )
  )

#ACEHURT1 
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_hurt = case_when(
      ACEHURT1 %in% c(1,7) ~ "0",
      ACEHURT1 %in% c(2,3) ~ "1",
      is.na(ACEHURT1) ~ "0"
    )
  )


#ACESWEAR
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping


g1data <- g1data%>%
  mutate(
    ace_swear = case_when(
      ACESWEAR %in% c(1,7) ~ "0",
      ACESWEAR %in% c(2,3) ~ "1",
      is.na(ACESWEAR) ~ "0"
    )
  )

#ACETOUCH 
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_touch = case_when(
      ACETOUCH %in% c(1,7) ~ "0",
      ACETOUCH %in% c(2,3) ~ "1",
      is.na(ACETOUCH) ~ "0"
    )
  )

#ACETTHEM
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_tthem = case_when(
      ACETTHEM %in% c(1,7) ~ "0",
      ACETTHEM %in% c(2,3) ~ "1",
      is.na(ACETTHEM) ~ "0"
    )
  )

#ACEHVSEX 
#1 = never 
#2 = once
#3 = more than once 
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_hvsex = case_when(
      ACEHVSEX %in% c(1,7) ~ "0",
      ACEHVSEX %in% c(2,3) ~ "1",
      is.na(ACEHVSEX) ~ "0"
    )
  )

#ACEADSAF 
#1 = never 
#2 = a little of the timne - coding as yes
#3 = some of the time - coding as yes
#4 - most of the time - coding as yes
#5 - all of the time - coding as yes
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_adsaf = case_when(
      ACEADSAF %in% c(1,7) ~ "0",
      ACEADSAF %in% c(2,3,4,5) ~ "1",
      is.na(ACEADSAF) ~ "0"
    )
  )


#ACEADNED 
#1 = never 
#2 = a little of the timne - coding as yes
#3 = some of the time - coding as yes
#4 - most of the time - coding as yes
#5 - all of the time - coding as yes
#7 = don't know not sure - coding as no
#9 = refused - dropping 
#blank - dropping

g1data <- g1data%>%
  mutate(
    ace_adned = case_when(
      ACEADNED %in% c(1,7) ~ "0",
      ACEADNED %in% c(2,3,4,5) ~ "1",
      is.na(ACEADNED) ~ "0"
    )
  )

## drops 

g1data <- g1data%>%
  filter(ace_deprs != 9,
         ace_drink != 9,
         ace_drug != 9,
         ace_prisn != 9, 
         ace_div != 9,
         ace_punch !=9,
         ace_hurt != 9,
         ace_swear != 9,
         ace_touch != 9, 
         ace_tthem != 9,
         ace_hvsex != 9,
         ace_adsaf != 9,
         ace_adned != 9
  )

###Ace count score 
g1data <- g1data %>%
  mutate(ace_count = rowSums(
    select(., ace_deprs:ace_adned) == 1))

g1data <- g1data %>%
  mutate(ace_count = as.numeric(ace_count))


g1data <- g1data %>%
  group_by(ace_count)%>%
   mutate(urate = (1 - mean(employ_dum, na.rm = TRUE)))%>%
  ungroup()




g1data <- g1data %>%
   mutate(avurate = (1 - mean(employ_dum, na.rm = TRUE)))

                            
#plot 
plot1 <- ggplot(data = g1data,
       mapping = aes(x= ace_count, y = urate))+
  geom_line()+
  geom_point()+ 
  scale_y_continuous(labels = scales::percent)+
    labs(title = "Adverse Childhood Experience Score and Unemployment Rate",
       subtitle = "Adult unemployment rates sharply increase as ACE score rises",
       caption = "Behavioral Risk Factor Surveillance System, 2021") +
  xlab("ACE Score") +
  ylab("Unemployment Rate") +
  theme_minimal() +
  theme(text = element_text(family = "Helvetica"),
        axis.text.x = element_text(
          angle = 45,
          vjust = 1,
          hjust = 1
        ),
        axis.title.y = element_text(
          angle = 90
          ),
        axis.title.x = element_text(
          hjust = .5,
          vjust = 1
        ))+
  geom_line(mapping = aes(y= avurate),
    color = "#00A087B2",
    linewidth = 1.25
  )

plot1 +
  annotate("text",
           x = 10,
           y = 0.17, 
           label = "(Sample Mean = 15.6%)",
           size = 3)
  
  

```


#Visualization 2
Along with ACEs, mental illness is strongly associated with poor employment and quality of life outcomes. Increased rates of depressive disorders can have significant impacts on a states' economy. In this graph, the average rate of reported diagnoses of depressive disorders are mapped by state. Oregon appears to have the highest diagnosis rates of depression, while North and Sotuh Dakota have the lowest. 
```{r}


g2data <- data



g2data<- g2data%>%
  mutate(state = `_STATE`)

#depression recode

g2data <- g2data %>%
  mutate(depr = case_when(
    ADDEPEV3 == 1 ~ 1,
    ADDEPEV3 == 2 ~ 0, 
    .default = NA
    )
  )
    


#loading greater 48 polygon data
statesdata <- tigris::states(cb = TRUE, progress_bar = FALSE) %>%
  filter(!STATEFP %in% c("78", "69", "66", "60", "72", "02", "15"))


g2data <- g2data%>%
  group_by(state)%>%
  mutate(mdepr = mean(depr, na.rm = TRUE)
  )

statesdata <- statesdata %>%
  mutate(STATEFP = as.double(STATEFP))

statesdata <- statesdata %>% 
  left_join(g2data, by = c("STATEFP" = "state")) %>%
  mutate(drate = if_else(!is.na(mdepr), mdepr, NA_real_))

statesdata <- statesdata %>%
  select("STATEFP", "STATENS", "AFFGEOID", "GEOID", "STUSPS",
             "NAME", "LSAD", "ALAND", "AWATER", "drate")


statesdata <- statesdata[!duplicated(statesdata$STATENS),]




statesdata %>%
  ggplot(mapping = aes(fill = drate)) +
  geom_sf() +
  scale_fill_continuous(labels = scales::percent,
    low = "#1097b8", 
    high = "#b83110") +
  theme_void() +  
  labs(title = "Depression by State, U.S. Adults, 2021",
       subtitle = "There are stark differences in depression diagnosis rates across the country",
       caption = "Behavioral Risk Factor Surveillance System, 2021") +
  theme(text = element_text(family = "Helvetica"),
        plot.title = element_text(vjust = 6),
        plot.subtitle = element_text(vjust = 5))+
  guides(fill = guide_colorbar(title = "Diagnosis Rate"))

  
```

#Visualization 3

This heatmap shows that poor, elderly Americans experience more days of poor physical health than any other income and age group. It is expected that the old have worse physical health than the young, comparing elderly Americans of higher and lower incomes reveals that poverty intensifies the negative effects of aging on health. The starkness of this inequality emphasizes the need for targeted aid to the elderly poor, a population that is uniquely vulnerable in the United States. The Social Security and Medicare systems are pre-existing mechanisms that could be applied towards reducing health inequities among our nations’ seniors.
```{r}
g3data <- data

g3data %>%
  filter(PHYSHLTH <= 30) %>%
  filter(INCOME3 <= 11) %>%
  filter(`_AGEG5YR` <= 13) %>%
  ggplot(mapping = aes(x = `_AGEG5YR`, y = INCOME3, fill = PHYSHLTH)) +
geom_tile() +
  coord_cartesian(xlim = c(1, 13), ylim = c(1, 11)) +
  scale_fill_continuous(
    low = "#1097b8", 
  high = "#b83110", 
  name = "Days of poor physical\nhealth per month") +
  scale_y_continuous(breaks = c(1:11),
                   labels = c("Less than $10,000",
                              "$10,000 - $15,000",
                              "$15,000 - $20,000",
                              "$20,000 - $25,000",
                              "$25,000 - $35,000",
                              "$35,000 - $50,000",
                              "$50,000 - $75,000",
                              "$75,000 - $100,000",
                              "$100,000 - $150,000",
                              "$150,000 - $200,000",
                              "$200,000 or more")) +
  scale_x_continuous(breaks = c(1:13),
                     labels = c("18 - 24",
                                "25 - 29",
                                "30 - 34",
                                "35 - 39",
                                "40 - 44",
                                "45 - 49",
                                "50 - 54",
                                "55 - 59",
                                "60 - 64",
                                "65 - 69",
                                "70 - 74",
                                "75 - 79",
                                "80 or older")) +
  labs(title = "Physical health among age and income groups, U.S. adults, 2021",
       subtitle = "Young, wealthy Americans report highest levels of\nphysical health, while poor, elderly Americans report the lowest",
       caption = "Behavioral Risk Factor Surveillance System, 2021") +
  xlab("Age") +
  ylab("Income") +
  theme_minimal() +
  theme(text = element_text(family = "Helvetica"),
        axis.text.x = element_text(
          angle = 45,
          vjust = 1,
          hjust = 1
        ),
        axis.title.y = element_text(
          angle = 0,
          hjust = 0,
          vjust = 1
        ),
        axis.title.x = element_text(
          hjust = .5,
          vjust = 1
        ))
  
```

#Visualization 4

This bar chart shows a negative association between income and mental health. Interestingly, the average reported days of mental distress increase fairly constantly as income decreases. This indicates that increased income leads to improved mental health outcomes regardless of ones’ current income. Boosting middle class incomes increases happiness about as much as increasing working class incomes, an effect that should be considered when crafting wage policy.

```{r}
g4data <- data

g4data <- g4data %>%
  filter(MENTHLTH <= 30) %>%
  group_by(INCOME3) %>%
  summarize(mean_MENTHLTH = mean(MENTHLTH))

g4data %>%
  filter(INCOME3 < 77) %>%
  ggplot(aes(x = INCOME3, 
             y = mean_MENTHLTH,
             fill = mean_MENTHLTH)) +
  geom_col() +
  scale_fill_continuous(low = "#1097b8", 
                        high = "#b83110",
                        guide = "none") +
  scale_x_continuous(breaks = c(1:11),
                   labels = c("Less than $10,000",
                              "$10,000 - $15,000",
                              "$15,000 - $20,000",
                              "$20,000 - $25,000",
                              "$25,000 - $35,000",
                              "$35,000 - $50,000",
                              "$50,000 - $75,000",
                              "$75,000 - $100,000",
                              "$100,000 - $150,000",
                              "$150,000 - $200,000",
                              "$200,000 or more")) +
  coord_flip() +
  labs(title = "Mental health by income groups, U.S. adults, 2021",
       subtitle = "Poor Americans report more days of mental distress than wealthy Americans",
       caption = "Behavioral Risk Factor Surveillance System, 2021",) +
  xlab("Annual income") +
  ylab("Average days of mental distress per month") +
  theme_minimal() +
  theme(text = element_text(family = "Helvetica"),
        axis.title.y = element_text(angle = 0, 
                                    hjust = 1,
                                     vjust = 1,
                                    )) 
  

```
